Statistical methods for relative quantification of post-translational modifications in global proteomics experiments

### 2021-02-01 main/MSstatsPTM_Main.tex
Updated main latex

\documentclass[11pt]{article}

@@

\author{
  LastName1, FirstName1\\
  \texttt{first1.last1@xxxxx.com}
  \and
  LastName2, FirstName2\\
  \texttt{first2.last2@xxxxx.com}
}

\begin{document}

\maketitle

@@

@@

Despite the important implications of PTMs in biological functions, there is a lack of general framework to summarize the available quantitative information from LC-MS data, to perform statistical inference, and to draw conclusions to characterize the quantitative properties of PTM in a statistically rigorous manner. Many investigations performed differential expression analysis of PTMs using two-sample t-test or its extensions. The approach takes as input intensities of individual features from modified peptides, or intensity ratios of modified and unmodified peptide features, and compares the mean abundance of a PTM site from one condition to another. Modifications of the t-test such as moderated t-test with limma were also proposed. While simple, the approach does not fully account for the sources of variations, and it is not directly applicable to experiments with complex designs, e.g., comparisons of multiple conditions, acquisition in multiple batches, etc. Isobar-PTM was developed for experiments with MS/MS quantitative strategies that employ isobaric labels such as tandem mass tags (TMT) and isobaric tag for relative and absolute quantification (iTRAQ).[Source] Isobar-PTM expresses MS measurements with a linear model and performs adjustment with respect to protein abundance using the difference between log-ratio of modified peptides in two channels and log-ratio of protein level. The modeling framework, however, is not applicable for either label-free workflows or experiments with complex designs. [Source]

@@

\section{Proposed Approach}

Figure 3 schematically illustrates a simplified version of the data structure resulting from a typical bottom-up experiment for quantitative analysis of PTMs, in which there are multiple layers of variation present. A PTM site is quantified with multiple spectral features, which vary in sequence (e.g., fully or partially cleaved peptides), ionization efficiency, charge states, etc. The number quantified features vary across replicate LC-MS/MS runs of the same sample, and across conditions. To perform adjustment with respect to protein abundance, features of unmodified peptides are used for the inference of underlying protein abundance. Typically, because of the enrichment step for PTMs, very few of those features are present in original LC-MS runs. For more accurate estimation of protein abundance, separate global proteomics data of unenriched samples are often acquired. As different levels of variability are present in the data, the log-intensities of the features for modified and unmodified peptides are modeled separately using two linear mixed models.

\subsection*{Statistical modeling and parameter estimation}

The proposed approach takes as input a list of log-transformed intensities of spectral features, identified and quantified across LC-MS runs. The features, which are precursor ions of modified or unmodified peptides, are used to characterize the identified PTM sites and proteins. For each PTM site, the feature log-intensities of the modified peptides spanning the site are expressed using a linear mixed model in consideration of the effects of condition, run, feature and interaction between run and feature.The model parameters are estimated using the split-plot approach as in MSstats[Source], where the feature log-intensities are first summarized into a single value per site per run in the subplot model, and the site-level summaries are then used for the inference of the PTM site abundance. In the site-level summarization, Tukey's median polish (TMP), a simple and robust procedure is applied to iteratively fit a two-way additive model with the effects of run and feature, which in turn summarizes the log-intensities for each site. [Source]  After the summarization, the inference of the PTM site abundance in each condition is carried out in account for the experimental design (Fig. 3a). Statistical modeling and quantification for global proteomics data are performed by the same procedure as for PTM data.

\subsection*{Detection of changes in PTMs}

@@

\paragraph*{Computer simulation: protein-level adjustment.} Differential intensity levels of modified peptides may be due to changes in modification, change in protein abundance, or both. The proposed approach adjusts the abundance with respect to unmodified peptides by combining the inference of modified and unmodified peptide abundances. Alternatively, two-sample t-test that takes as input the ratio between modified and unmodified peptide intensities (difference on log scale) is commonly applied for the same purpose. In real experiments, multiple inter-related conditions are often compared together. Whereas t-test uses measurements from the two conditions being compared, the proposed approach leverages measurements in all conditions for the inference of underlying abundance. We evaluated the impact with such adjustment by computer simulation, considering the following factors: with/without protein-level changes, effect size, number of replicates and number of conditions. Details of the simulation can be found in Supplementary Sec. S4.1. 
In this simulation, when there was no change in protein abundance, all the considered approaches well calibrated the Type I error rate (Fig. 4a). However, when the modification changes were entirely due to the changes in protein abundance across conditions, analysis without accounting for the protein-level changes resulted in off-target, high false positive rates (Fig. 4b). As shown in Fig. 4c-e, in detection of systematic changes in PTM abundance, the proposed approach improved statistical power with small sample sizes in almost all the considered scenarios. The advantage by using the proposed approach over t-test became more profound in the presence of missing data (Fig. 4d-e). In cases with small sample sizes (e.g., 2 replicates) as shown in Fig. 4d, performance by t-test decreased dramatically with one missing value, and increasing the log-fold change to 2 did not effectively reduce the negative impact. Two-sample t-test only used data within the groups of interest while ignoring the rest of the data. Consequently, it gave similar performance across cases with different number of conditions. In contrast, the proposed approach leveraged all available information, which resulted in improved power with increased number of conditions (see for example, the improved performance in Fig. 4e [with 4 conditions] over Fig. 4d [with 2 conditions]). Further results are provided in Supplementary Sec. S4.1.

@@

\subsubsection*{Spike-in experiment [Need to write results of this experiment]}

@@

@@

\section{Figure legends}

@@

>>>>

\documentclass[11pt]{article}
%\documentclass[11pt, twocolumn]{article}

\usepackage[colorinlistoftodos]{todonotes}
\usepackage{graphicx}
\graphicspath{ {./images/} }

@@

%\author{
%  LastName1, FirstName1\\
%  \texttt{first1.last1@xxxxx.com}
%  \and
%  LastName2, FirstName2\\
%  \texttt{first2.last2@xxxxx.com}
%}
\date{}
\begin{document}

\maketitle
\todo{Add Author list}

@@

The scientific community widely utilizes mass spectrometry (MS)-based proteomics to quantify the abundance of proteins and their post-translational modifications (PTMs). Experiments targeting PTMs face several specific challenges. These include the low abundance of modified proteo-forms, few representative peptides that span modification sites, and convolution with abundance changes in the overall protein expression. Due to these challenges, a robust approach to estimate relative systematic changes in PTMs should combine information pertaining to PTM sites over several peptides, replicates in multiple conditions, and consider sources of confounding and variation present in the experiment. We propose a general statistical model and workflow that is both reproducible and comprehensive. The model measures PTM and protein abundance by summarizing intensities through Tukey’s median polish method. Then a model based on the family of linear mixed-effects models is fit. This model is automatically adjusted to the specific experimental design. Finally, the PTM abundances are adjusted to remove variance from changes in the overall protein. We implement this model in the free and open-source R package MSstatsPTM.
\todo{Used overview from USHUPO poster}

@@

Despite the important implications of PTMs in biological functions, there is a lack of general framework to summarize the available quantitative information from LC-MS data, to perform statistical inference, and to draw conclusions to characterize the quantitative properties of PTM in a statistically rigorous manner. Many investigations performed differential expression analysis of PTMs using two-sample t-test or its extensions. The approach takes as input intensities of individual features from modified peptides, or intensity ratios of modified and unmodified peptide features, and compares the mean abundance of a PTM site from one condition to another. Modifications of the t-test such as moderated t-test with limma were also proposed. While simple, the approach does not fully account for the sources of variations, and it is not directly applicable to experiments with complex designs, e.g., comparisons of multiple conditions, acquisition in multiple batches, etc. Isobar-PTM was developed for experiments with MS/MS quantitative strategies that employ isobaric labels such as tandem mass tags (TMT) and isobaric tag for relative and absolute quantification (iTRAQ).[Source] Isobar-PTM expresses MS measurements with a linear model and performs adjustment with respect to protein abundance using the difference between log-ratio of modified peptides in two channels and log-ratio of protein level. The modeling framework, however, is not applicable for either label-free workflows or experiments with complex designs.
\todo{Add Source. Compare this to the TMT part of MSstatsPTM}

@@

\section{Proposed Approach}
\begin{center}
\includegraphics[scale=.8]{fig3}
\end{center}
Figure 3 schematically illustrates a simplified version of the data structure resulting from a typical bottom-up experiment for quantitative analysis of PTMs, in which there are multiple layers of variation present. A PTM site is quantified with multiple spectral features, which vary in sequence (e.g., fully or partially cleaved peptides), ionization efficiency, charge states, etc. The number quantified features vary across replicate LC-MS/MS runs of the same sample, and across conditions. To perform adjustment with respect to protein abundance, features of unmodified peptides are used for the inference of underlying protein abundance. Typically, because of the enrichment step for PTMs, very few of those features are present in original LC-MS runs. For more accurate estimation of protein abundance, separate global proteomics data of unenriched samples are often acquired. As different levels of variability are present in the data, the log-intensities of the features for modified and unmodified peptides are modeled separately using two linear mixed models.

\subsection*{Statistical modeling and parameter estimation}
\todo{Add in TMT approach}
The proposed approach takes as input a list of log-transformed intensities of spectral features, identified and quantified across LC-MS runs. The features, which are precursor ions of modified or unmodified peptides, are used to characterize the identified PTM sites and proteins. For each PTM site, the feature log-intensities of the modified peptides spanning the site are expressed using a linear mixed model in consideration of the effects of condition, run, feature and interaction between run and feature.The model parameters are estimated using the split-plot approach as in MSstats[Source], where the feature log-intensities are first summarized into a single value per site per run in the subplot model, and the site-level summaries are then used for the inference of the PTM site abundance. In the site-level summarization, Tukey's median polish (TMP), a simple and robust procedure is applied to iteratively fit a two-way additive model with the effects of run and feature, which in turn summarizes the log-intensities for each site. [Source]  After the summarization, the inference of the PTM site abundance in each condition is carried out in account for the experimental design (Fig. 3a). Statistical modeling and quantification for global proteomics data are performed by the same procedure as for PTM data.  
\todo{Add Sources}
\subsection*{Detection of changes in PTMs}

@@

\paragraph*{Computer simulation: protein-level adjustment.}

Differential intensity levels of modified peptides may be due to changes in modification, change in protein abundance, or both. The proposed approach adjusts the abundance with respect to unmodified peptides by combining the inference of modified and unmodified peptide abundances. Alternatively, two-sample t-test that takes as input the ratio between modified and unmodified peptide intensities (difference on log scale) is commonly applied for the same purpose. In real experiments, multiple inter-related conditions are often compared together. Whereas t-test uses measurements from the two conditions being compared, the proposed approach leverages measurements in all conditions for the inference of underlying abundance. We evaluated the impact with such adjustment by computer simulation, considering the following factors: with/without protein-level changes, effect size, number of replicates and number of conditions. Details of the simulation can be found in Supplementary Sec. S4.1. 
In this simulation, when there was no change in protein abundance, all the considered approaches well calibrated the Type I error rate (Fig. 4a). However, when the modification changes were entirely due to the changes in protein abundance across conditions, analysis without accounting for the protein-level changes resulted in off-target, high false positive rates (Fig. 4b). As shown in Fig. 4c-e, in detection of systematic changes in PTM abundance, the proposed approach improved statistical power with small sample sizes in almost all the considered scenarios. The advantage by using the proposed approach over t-test became more profound in the presence of missing data (Fig. 4d-e). In cases with small sample sizes (e.g., 2 replicates) as shown in Fig. 4d, performance by t-test decreased dramatically with one missing value, and increasing the log-fold change to 2 did not effectively reduce the negative impact. Two-sample t-test only used data within the groups of interest while ignoring the rest of the data. Consequently, it gave similar performance across cases with different number of conditions. In contrast, the proposed approach leveraged all available information, which resulted in improved power with increased number of conditions (see for example, the improved performance in Fig. 4e [with 4 conditions] over Fig. 4d [with 2 conditions]). Further results are provided in Supplementary Sec. S4.1.

@@

\subsubsection*{Spike-in experiment}
\todo{Need to write results of this experiment}
\todo{Devon: I don't know what these Label-Free experiments are or how to analyze them. Meeting with Tsung-Heng on 2/5 to review}

@@

\todo{Add Shigella for TMT side}

@@

\section{Figure legends}
\todo{Unclear where Tsung-Heng meant these plots to go. Need to add them into correct sections}

@@

\todo{Add sources section}
